The Design and Implementation of the Synchronous Language \CEU

### 2016-04-11 paper.tex
...

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and device drivers.
%
We evaluated the expressiveness of \CEU in comparison to hand-crafted 
event-driven code in \C and attested a reduction in source code size (around 
25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes and can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
in comparison to a hand-crafted event-driven code base%
\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}
in \emph{nesC} (a \C variant)~\cite{wsn.nesc}.
%
Figure~\ref{fig.eval} compares source size (\emph{number of tokens}), binary 
size (\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
network protocols and a radio driver.
The small overhead in resource usage shows that the gains in productivity and 
safety with \CEU make it a viable alternative to \C in the context of 
constrained embedded systems.
%
Unfortunately, most real applications in Esterel seem to be closed source.
In a future work, we plan to cooperate with groups that have more experience 
and access to real code bases in Esterel for a comprehensive comparison with 
\CEU.
% TODO-no:
%Still, considering \C as an ``absolute benchmark'', we consider the small 
%overhead .  solid results

@@

In the subsections that follow, we discuss \CEU-specific implementation 
details:
temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
single-threaded dispatching (Section~\ref{sec.impl.single}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

For each node representing a statement in the program AST, we keep the set of 
input events $I$ (\emph{incoming}) that can lead to the execution of the node, 
and also the set of input events $O$ (\emph{outgoing}) that can terminate the 
node.
As an example, for the single-statement program \code{await~A}, we have 
$I=\{boot\}$ and $O=\{A\}$.

A node inherits the set $I$ from its immediate parent and calculates $O$ 
according to its type, as follows:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, \C calls, and declarations 
simply reproduce $O=I$, as they do not await;
%
\item An \code{await E} statement, where \code{E} is an external input event,
      has $O=\{E\}$ (see also internal events below).
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never proceeds to the statement immediately following it (see 
also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$, except for $n=1$ which inherits $I$ from the parent node.
The set $O$ for the whole node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes the output of its body on its own $I$ ($I=I 
\cup O_{body}$), as the loop is also reached from its own body.
The union of all $O$ from nested \code{break} statements forms the set $O$ for 
a \code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$, where \code{true} and 
\code{false} are the two clauses of the \code{if}.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\item For internal events, an \code{await} awakes from any input that leads to 
    any matching \code{emit} in a trail in parallel:
    \begin{itemize}
    \item An \code{await e} has $O=I_{e1} \cup ... \cup I_{eN}$,
        where $e1...eN$ are \code{emit e} statements in trails in parallel.
    \item An \code{emit e} terminates in the same reaction, having $O=I$.
    \end{itemize}
\end{itemize}

@@

The code in Figure~\ref{lst.impl.ast}.a has its corresponding $AST$ and sets 
$I$ and $O$ in Figure~\ref{lst.impl.ast}.b.
The event $.$ (dot) represents the ``boot'' reaction.
The assignments to \code{y} in parallel (ln. \ax:5,8) have an empty 
intersection of $I$ (ln. \bx:6,9), hence, they do not conflict.
Note that, although the accesses to \code{y} in sequence (ln. \ax:5,11)
do have an intersection (ln. \bx:6,11), they are not in parallel branches and 
are also safe.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\centering\small{\ax} A program in \CEU...
\end{minipage}
%
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\centering\small{\bx} ...with corresponding sets $I$ and $O$.
\end{minipage}
%
%\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

\CEU favors a fine-grained use of trails, being common to use trails that await 
a single event and terminate.
For this reason, \CEU does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{union}, while blocks in parallel, 
in a \code{struct}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at any given time.
A position in the memory may hold different data (with variable sizes) during 
runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with its corresponding 
memory layout.
%
Each variable is assigned a unique $id$ so that variables with the same name 
can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
\ax:3).
%
The \code{do-end} blocks and \code{par/and} in sequence
(ln. \ax:2--4,5--7,8--12) are packed in a \code{union} (ln. \bx:2,3,4--7), 
given that their variables cannot be in scope at the same time, e.g., 
\code{MEM.a\_1} and \code{MEM.b\_2} (ln.  \bx:2,3) can safely share the same 
memory slot.
%
The example also illustrates the presence of runtime flags (ln. \bx:4--7) 
related to the \code{par/and} termination (ln. \ax:8--12), which also reside in 
reusable slots in the static memory.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\centering\small{\ax} A program in \CEU...
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;

@@

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds an associated numeric stack level and can only 
execute when scheduler runtime reaches that level.
%
A \code{FINALIZE} trail represents a hanged finalization block which is
scheduled only  when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the minimum sequence number (\emph{seqno}) in which it can awake 
(for \emph{delayed awaits}).

to prevent from awaking in the same reaction it is reached.
%
In concrete terms, a trail is represented by the following \code{struct}:

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two 
segments and requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
Since the scheduler traverses all trails on every reaction, it adjusts all 
\code{seqno} to avoid overflows (actually, 2 bits to hold the \code{seqno} is 
already enough).
%
The size of \code{stack\_t} depends on the maximum depth of nested emissions 
but is bounded to the maximum number of trails:
in the worst case, a trail emits an event that awakes another trail, which 
emits an event that awakes another trail, and so on;
the last trail cannot awake any trail, because they are all hanged in the 
\code{STACKED} state.

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with memory overhead of only 2\% in comparison 
to the original single-threaded version in \C~\cite{ceu.sensys13}.

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in \ax, the compiler extracts the entry points and associated 
trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0}
(ln. \ax:7 and \bx:3).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail in two segments with 
associated entry points.
%
The entry points translate to an \code{enum} in the generated code (ln.  
\bx:1--10).
The state of trails translates to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. \bx:12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. \bx:13), while all others are set to \code{INACTIVE} (ln. \bx:14).

@@

The code in \cx dispatches a trail segment according to the current label to 
execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and} (ln. \ax:4), it stacks 
\code{TRAIL-1} (ln. \cx:4--7) and proceeds to the code in \code{TRAIL-0} (ln.  
\cx:9--14) to respect lexical execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
Then, \code{dispatch} takes \code{TRAIL-1} and executes its registered label 
\code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to await 
\code{EVT\_A} and also halts.

@@

The implementation of \CEU dispatches active trails sequentially in a single 
thread, taking no advantage of multi-core CPUs.
%
This decision comes not only from the fact that \CEU targets constrained 
embedded systems with a single CPU, but also because \CEU imposes deterministic 
execution for intra-reaction statements.

Note that, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
analysis of \CEU infers precisely trails that are concurrent yet do not share 
resources.
Hence, these non-conflicting trails could potentially execute with real 
parallelism in multiple cores.
%
However, our experiments with multi-core execution are actually slower than 
single-core execution in the same system.
Considering that we use \CEU primarily in control-dominated applications, this 
result is not surprising and also appears in related 
work~\cite{esterel.multicore.1,esterel.multicore.2}.
One reason is due to continuous fork-and-rejoin overhead from small reactions.
Another reason is due to excessive locality of data from stackless trails 
sharing contiguous static memory.

@@

As a reactive language, the execution of programs in \CEU is guided entirely by 
the occurrence of external input events.
%
The binding for a specific platform (environment) calls hook functions in the 
API of the \CEU runtime whenever an external event occurs.
%
However, to preserve the sequential/discrete semantics of time in \CEU, these 
calls must never interleave or parallelize execution.

@@

The main difference between the standard \C back end and the Terra \VM is the 
\emph{code generation phase}, which here outputs assembly instructions for the 
\VM, instead of statements in \C.
%
To reduce the memory footprint of applications, the \VM includes special 
instructions for complex and recurrent operations from the runtime of \CEU, 
such as for handling events and trails.

>>>>

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and device drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes and can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
in comparison to a code base%
\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}
in \emph{nesC} (a \C variant)~\cite{wsn.nesc}.
%
Figure~\ref{fig.eval} compares source size (\emph{number of tokens}), binary 
size (\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
network protocols and a radio driver.
The small overhead in resource usage shows that the gains in productivity and 
safety with \CEU make it a viable alternative to \C in the context of 
constrained embedded systems.
%
Unfortunately, most real applications in Esterel seem to be closed source.
In a future work, we plan to cooperate with groups that have more experience 
and access to real code bases in Esterel for a comprehensive comparison with 
\CEU.
% TODO-no:
%Still, considering \C as an ``absolute benchmark'', we consider the small 
%overhead .  solid results

@@

In the subsections that follow, we discuss implementation details specific to 
\CEU:
temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
single-threaded dispatching (Section~\ref{sec.impl.single}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

For each node representing a statement in the program AST, we keep the set of 
input events $I$ (\emph{incoming}) that can start the execution of the node, 
and also the set of input events $O$ (\emph{outgoing}) that can terminate the 
node.
As an example, for the single-statement program \code{await~A}, we have 
$I=\{boot\}$ and $O=\{A\}$.

A node inherits the set $I$ from its immediate parent and calculates $O$ 
according to its type, as follows:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, \C calls, and declarations 
simply reproduce $O=I$, as they do not await;
%
\item An \code{await E} statement, where \code{E} is an external input event,
      has $O=\{E\}$ (see also internal events below).
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never proceeds to the statement immediately following it (see 
also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$, except for $n=1$ (which inherits $I$ from the parent node).
The set $O$ for the whole node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes the output of its body on its own $I$ ($I=I 
\cup O_{body}$), as the loop is also reached from its own body.
The union of all $O$ from nested \code{break} statements forms the set $O$ for 
a \code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$, where \code{true} and 
\code{false} are the two \code{if} branches.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\item For internal events, an \code{await} awakes from any input that leads to 
    any matching \code{emit} in a trail in parallel:
    \begin{itemize}
    \item An \code{await e} has $O=I_{e1} \cup ... \cup I_{eN}$,
        where $e1...eN$ are \code{emit e} statements in trails in parallel.
    \item An \code{emit e} terminates in the same reaction, having $O=I$.
    \end{itemize}
\end{itemize}

@@

The code in Figure~\ref{lst.impl.ast}.a has its corresponding $AST$ and sets 
$I$ and $O$ in Figure~\ref{lst.impl.ast}.b.
%The event $.$ (dot) represents the ``boot'' reaction.
The assignments to \code{y} in parallel (ln. \ax:5,8) have an empty 
intersection of $I$ (ln. \bx:6,9), hence, they do not conflict.
Note that, although the accesses to \code{y} in sequence (ln. \ax:5,11)
do have an intersection (ln. \bx:6,11), they are not in parallel branches and 
are also safe.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\centering\small{\ax} A program in \CEU...
\end{minipage}
%
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={boot} O={A}
    Dcl_y I={boot} O={boot}
    ParOr I={boot} O={A,B}
        Stmts I={boot} O={A}
            Await_A I={boot} O={A}
            Set_y I={A} O={A}
        Stmts I={boot} O={B}
            Await_B I={boot} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\centering\small{\bx} ...with corresponding sets $I$ and $O$.
\end{minipage}
%
%\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

\CEU favors a fine-grained use of trails, being common to use trails that await 
a single event and terminate.
For this reason, \CEU does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm}: memory for 
blocks in sequence are packed in a \code{union}, while blocks in parallel, in a 
\code{struct}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at any given time.
A position in the memory may hold different data (with variable sizes) during 
runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with its corresponding 
memory layout.
%
%Each variable is assigned a unique $id$ so that variables with the same name 
%can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
%\ax:3).
%
The \code{do-end} blocks and \code{par/and} in sequence
(ln. \ax:2--4,5--7,8--12) are packed in a \code{union} (ln. \bx:2,3,4--7), 
given that their variables cannot be in scope at the same time, e.g., \code{a} 
and \code{b} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags (ln. \bx:4--7) 
related to the \code{par/and} termination (ln. \ax:8--12), which also reside in 
reusable slots in the static memory.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\centering\small{\ax} A program in \CEU...
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
union {             // sequence
    int a;          //   do_1
    int b;          //   do_2
    struct {        //   par/and
        int _and_1: 1;
        int _and_2: 1;
    };
} MEM ;

@@

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds an associated numeric stack level and can only 
execute when scheduler runtime drops to that level.
%
A \code{FINALIZE} trail represents a hanged finalization block which is
scheduled only  when its corresponding block goes out of scope.
%
A trail waiting for an event stays in that event, also holding the minimum 
sequence reaction number (\emph{seqno}) in which it can awake (to respect
\emph{delayed awaits}).
%
In concrete terms, a trail is represented by the following \code{struct}:

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two 
segments and requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The fields \code{seqno} requires only 2 bits because the scheduler adjusts them 
while traversing all trails.
%
The size of \code{stack\_t} depends on the maximum depth of nested emissions 
but is bounded to the maximum number of trails:
in the worst case, a trail emits an event that awakes another trail, which 
emits an event that awakes another trail, and so on;
the last trail cannot awake any other trail, because they are all hanged in the 
\code{STACKED} state.

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution but has a memory overhead of only 2\% in 
comparison to the original single-threaded version in \C~\cite{ceu.sensys13}.

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in \ax, the compiler extracts the entry points and associated 
trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0}
(ln. \ax:7 and \bx:3).
%
For each yielding statement (e.g., \code{emit}, \code{await}, \code{par/and}, 
etc.), the compiler splits the trail in two segments with associated entry 
points.
%
The entry points translate to an \code{enum} in the generated code (ln.  
\bx:1--10).
The state of trails translates to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. \bx:12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. \bx:13), while all others are set to \code{INACTIVE} (ln. \bx:14).

@@

The code in \cx dispatches a trail segment according to the current label to 
execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0} 
(ln.~3--14).
%
When the \code{Main} label reaches the \code{par/and} (ln. \ax:4), it first 
stacks \code{TRAIL-1} (ln. \cx:4--7) and then executes the \code{await~e} 
(ln.~\ax:6) in \code{TRAIL-0} (ln.~\cx:9--14), respecting lexical execution 
order.
%
The dispatcher sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
Then, it switches to \code{TRAIL-1} and executes label \code{And\_sub\_2} 
(ln.~\cx:6,16--21), which sets \code{TRAIL-1} to await \code{EVT\_A} and also 
halts.

@@

The implementation of \CEU dispatches active trails sequentially in a single 
thread, taking no advantage of multi-core CPUs.
%
This decision comes not only from the fact that \CEU targets constrained with 
single-CPU embedded systems, but also because \CEU imposes deterministic 
execution for intra-reaction statements.

Note that, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
analysis of \CEU infers precisely trails that are concurrent and yet do not 
share resources.
Hence, these non-conflicting trails could potentially execute with real 
parallelism in multiple cores.
%
However, our experiments with multi-core execution are actually slower than 
single-core execution in the same system.
Considering that we use \CEU primarily in control-dominated applications, this 
result is not surprising and also appears in related 
work~\cite{esterel.multicore.1,esterel.multicore.2}.
One reason is due to overhead from continuous fork-and-rejoin in small 
reactions.
Another reason is due to contention form excessive locality of data in 
stackless trails sharing contiguous static memory.

@@

As a reactive language, the execution of programs in \CEU is guided entirely by 
the occurrence of external input events.
%
The binding for a specific platform (environment) calls hook functions in the 
API of the \CEU runtime whenever an external event occurs.
%
These calls must never interleave or parallelize execution in order to preserve 
the sequential/discrete notion of time in \CEU.

@@

The main difference between the standard \C back end and the Terra \VM is the 
\emph{code generation phase}, which here outputs assembly instructions for the 
\VM (instead of statements in \C).
%
To reduce the memory footprint of applications, the \VM includes special 
instructions for complex and recurrent operations from the runtime of \CEU, 
such as for handling events and trails.
